%% =============================================================================
%% DEFINING CHAT-FIRST ANALYTICS
%% =============================================================================

\section{Background and design principles} \label{sec:defining}

The introduction identified three concerns with using language models for
data analysis: correctness, privacy, and reproducibility. This section
examines these shortcomings more carefully, then states the three
principles that define chat-first analytics.

\subsection{How LLM-assisted analysis goes wrong}

The appeal of typing ``regress wages on education with robust standard
errors'' into ChatGPT is obvious. The problems are less visible.

First, \emph{correctness and auditability.} LLM-generated code often runs
without error but computes the wrong quantity---for instance, requesting
``clustered standard errors'' might yield heteroskedasticity-robust
errors instead, a panel model might omit entity fixed effects, or a
difference-in-differences specification might use the wrong interaction
term. \citet{Ludwig+Mullainathan+Rambachan:2025} provide a systematic
framework for reasoning about these risks: the code executes, returns
plausible-looking output, and the user has no way to detect the error
without the expertise that would have let them write the code
themselves. The problem is
compounded by opacity: when a system generates 30 lines of
\proglang{Python} to answer a statistical question, understanding what
was computed requires reading and auditing those 30 lines. The user
who could not write the code cannot audit it either. This extends
beyond research: enterprise text-to-SQL deployments struggle with the
same issue---ambiguous questions admit multiple valid interpretations,
and users cannot verify which was
executed~\citep{Renggli+Ilyas+Rekatsinas:2025}.

Second, \emph{privacy.} Code-generation systems require the data.
When a user asks ChatGPT to analyze a dataset, the data are uploaded
to a third-party server with unclear retention policies. For survey
responses, health records, personnel data, or any dataset subject to
privacy regulations, this may be unacceptable. Even when providers
promise not to retain data, the transmission itself may violate
institutional or regulatory requirements.

Third, \emph{reproducibility.} The same prompt sent to the same model
on different days may produce different code. Models are updated,
fine-tuned, and retrained. The generated \proglang{Python} script that
worked in January may not be what the model produces in March.
Conversational sessions are ephemeral---unless the user manually saves
the generated code, the analysis cannot be reproduced. This
contradicts a basic requirement of scientific work.

These three problems are distinct but related. Silent errors arise
because the language model controls both \emph{what} to compute and
\emph{how} to compute it, and the computational logic is
opaque---buried in generated code rather than invoked from a known,
inspectable implementation. Privacy violations arise because
computation and data must be co-located on the provider's servers.
Reproducibility failures arise because the generated code is a
function of the model's current state.
Systematic error rates for LLM-generated statistical code are not
yet well established: \citet{Ludwig+Mullainathan+Rambachan:2025}
document the \emph{types} of errors that arise, but base rates
depend on the model, the prompt, and the statistical method. The
motivation for the architecture described below rests on the
existence and nature of these shortcomings, not on their precise
frequency.

\subsection{Three principles} \label{sec:principles}

Chat-first data analytics addresses these problems through three
principles. ``Chat-first'' means that natural language conversation
is the designed-for entry point: tool schemas are annotated for LLM
consumption, error messages are phrased for conversational context,
and default parameters are chosen so that typical requests need
minimal specification. It does not mean chat-only---\pkg{prompt2analytics}
also provides a CLI that invokes the same methods directly
(Section~\ref{sec:tools}), and every tool call maps to a public
\proglang{Rust} function usable without any LLM---but the
conversational interface is the primary design target. The first
two principles are definitional; the third is a design choice that we
consider essential.

\subsubsection{Principle 1: Separate interpretation from computation}

The language model decides \emph{what} to compute; validated software
decides \emph{how}. When a user asks to ``estimate the effect of
education on wages, controlling for experience, with robust standard
errors,'' the language model's job is to determine that this requires
OLS with HC1 standard errors, identify the column names, and generate
a structured tool call. The actual matrix algebra---forming
$(\bm{X}'\bm{X})^{-1}\bm{X}'\bm{y}$, computing the
HC1 sandwich estimator---is performed by pre-implemented code that
the language model cannot modify.

This division of labor plays to each component's strengths. Language
models are good at understanding intent, resolving ambiguity, and
mapping natural language to structured specifications. They can also
generate correct code much of the time---but offer no guarantee that
they have done so, and no mechanism for the user to verify correctness
short of reading the code. Separation sidesteps this problem entirely:
the model never produces code at all.
Wrong parameter extraction remains possible (Section~\ref{sec:evaluation}),
but the error surface is smaller: the model selects from a fixed schema
rather than generating arbitrary code.

The idea has precedent in text-to-SQL systems~\citep{Yu+etal:2018,
Li+etal:2024}, where a language model translates natural language into
structured database queries executed by a SQL engine. We extend this
pattern from data retrieval to statistical computation: from
text-to-SQL to text-to-statistics.

\subsubsection{Principle 2: Pre-validate methods, not generated code}

Correctness is established once, through systematic validation,
rather than hoped for anew with each request. When
\pkg{prompt2analytics} runs OLS, it uses the same \proglang{Rust}
function every time---a function validated against \proglang{R}'s
\fct{lm} on the Longley dataset to $10^{-8}$ precision
(Section~\ref{sec:evaluation}). Of course, an LLM-generated script
would also call a trusted function like \fct{lm}---the underlying
library is not the problem. The problem is the ad-hoc code
\emph{around} the library call: selecting variables, reshaping data,
passing options, and post-processing output. In the code-generation
approach, this glue code is generated anew each time and must be
audited each time. In the pre-validated approach, it is part of the
tested implementation and never changes.

This principle also solves the reproducibility problem. The same
method invocation with the same data produces identical results,
regardless of when it is run or which language model is used. The
stochastic element of the language model affects only interpretation
(which tool to call, how to explain the output), never computation.
Two users who phrase the same analysis differently may get different
tool calls---but once the tool is selected and parameters specified,
the result is deterministic.

An obvious question is why not wrap existing \proglang{R} or
\proglang{Python} packages rather than re-implement methods in
\proglang{Rust}. Three practical considerations favor a unified
implementation. First, \emph{standardization}: a single library can
expose every method through a uniform API---the same column-based
calling convention, the same error types, the same result
structures---which maps cleanly to tool schemas for language model
consumption. Wrapping heterogeneous packages (each with its own
formula syntax, option names, and output format) would require
per-package glue code that is itself a source of bugs. Second,
\emph{dependency management}: covering 200+ methods via existing
packages would require dozens of package dependencies with their own
version constraints, system library requirements, and breaking
changes across releases---a maintenance burden that grows
combinatorially. A self-contained implementation eliminates this
entirely. Third, \emph{deployment}: a single compiled binary with no
runtime dependencies is straightforward to distribute and run on any
platform, which matters for the local-first deployment model
described in Principle~3. None of this implies that existing packages
are inadequate---we validate against them precisely because they are
the reference standard (Section~\ref{sec:evaluation}).

\subsubsection{Principle 3: Keep data local; transmit only structure}

The language model receives column names, data types, and summary
statistics---not the data itself. A dataset of 100,000 salary records
is represented to the model as metadata: ``columns: employee\_id (int),
salary (float, range 28000--185000), department (string, 12 unique
values), years\_experience (float, range 0.5--35.2).'' Statistical
computations execute on the user's machine; only aggregated results
(coefficients, standard errors, test statistics) pass through the
language model for interpretation.

This is not definitionally required for chat-first analytics---a
cloud-based system satisfying Principles~1 and~2 would still qualify.
But we consider data locality essential for the use cases that
motivate this work. Researchers analyzing survey data, firms analyzing
personnel records, and analysts working under privacy regulations need
architectures where data privacy is guaranteed by design, not
promised by policy.

For complete privacy, the language model itself can run locally.
Open-source models deployed via Ollama~\citep{Ollama} eliminate all
external network communication. Our evaluation
(Section~\ref{sec:evaluation}) shows that even a 3-billion-parameter
model achieves over 80\% adequate rate, making fully local
deployment practical on consumer hardware. Given the rapid
improvement in small open-source models---each generation has brought
substantial gains on structured tasks like tool selection and
parameter extraction---this lower bound is likely to rise
considerably in the near term.

Together, the three principles address the three concerns identified
above. Separation and pre-validation eliminate silent computational
errors and improve auditability: the model never generates code, and
the user inspects a structured tool call rather than 30 lines of
generated script. Pre-validation also ensures reproducibility, because
the same method invocation with the same data always produces the same
result. Data locality protects privacy by keeping raw data on the
user's machine.

\subsection{Relationship to existing ideas}

Chat-first analytics applies established techniques to a new domain.
Tool-augmented language models~\citep{Schick+etal:2023, Yao+etal:2023}
can reliably select and invoke external tools when provided with
appropriate schemas~\citep{Qu+etal:2025, Liu+etal:2025:ToolACE}. The
Model Context Protocol (MCP)~\citep{MCP:2024} standardizes how
language models discover and invoke tools, providing the communication
layer between interpretation and computation.%
\footnote{MCP focuses on model-to-tool integration. Complementary
protocols address agent-to-agent communication: Google's Agent2Agent
(A2A) protocol~\citep{A2A:2025} and IBM's Agent Communication
Protocol (ACP)~\citep{ACP:2025}, which merged under Linux Foundation
governance in 2025. For single-agent tool invocation---the pattern
used here---MCP is the dominant standard, with adoption by OpenAI,
Microsoft, and Google alongside Anthropic.}

The interface innovation has precedents in four strands of work.
First, \emph{formula notation and domain-specific languages}:
the Wilkinson-Rogers formula notation~\citep{Wilkinson+Rogers:1973}
simplified model specification in the 1970s;
\citet{Chambers+Hastie:1992} extended it in \proglang{S} and
\proglang{R}~\citep{Chambers:1998}; and probabilistic programming
languages like Stan~\citep{Carpenter+etal:2017} apply the DSL
pattern to Bayesian inference. Chat-first analytics replaces
formal notation with natural language, trading precision for
accessibility.
Second, \emph{natural language interfaces to data systems}:
LUNAR~\citep{Woods:1973} demonstrated feasibility for database
querying in the 1970s, and the text-to-SQL paradigm---driven by
benchmarks like Spider~\citep{Yu+etal:2018}---has produced
substantial improvements in translating natural language to
structured queries~\citep{Li+etal:2024}. We extend this pattern
from data retrieval to statistical computation.
Third, \emph{automated machine learning}: Auto-sklearn~\citep{Feurer+etal:2015}
and H2O AutoML~\citep{LeDell+Poirier:2020} automate model selection
through algorithmic search, and recent work integrates LLMs into
the pipeline~\citep{Hollmann+etal:2023}. AutoML optimizes for
prediction; chat-first analytics interprets user intent across
prediction, causal inference, hypothesis testing, and description.
Fourth, \emph{computational econometrics} has established
standards for numerical accuracy~\citep{McCullough:1998,
McCullough+Vinod:1999, Davidson+MacKinnon:2004}, which motivate
the validation approach described in Section~\ref{sec:evaluation}.

Several systems have explored LLM-assisted data analysis, but with
a code-generation architecture. \citet{Hong+etal:2024} present a
``Data Interpreter'' that generates and executes \proglang{Python}
code. \citet{Chen+etal:2025} develop an ``Econometrics AI Agent''
that augments code generation with domain knowledge.
\citet{Sun+etal:2025:LAMBDA} describe an agent that orchestrates
generated code across analysis stages. All three generate code;
none pre-validate implementations or enforce data locality.
Table~\ref{tab:system-comparison} summarizes these differences.

Hybrid designs are possible, and the boundaries are not sharp.
A code-generation system that produces a short script calling
\proglang{R}'s \fct{felm} or \proglang{Python}'s
\fct{linearmodels.PanelOLS} already delegates computation to a tested
library---the remaining vulnerability is in the glue code (variable
selection, data reshaping, option passing), not in the computation
itself. A hybrid system could route standard requests to pre-validated
tools while falling back to code generation for bespoke analyses, or
it could verify generated scripts through static analysis or human
review before execution. \pkg{prompt2analytics} deliberately occupies
one end of this spectrum: all computation is pre-validated, with no
code generation at all. This sacrifices flexibility---analyses outside
the method library are unavailable, custom data transformations beyond
the built-in munging tools require the CLI or direct \proglang{Rust}
code, and ad-hoc visualizations not covered by the visualization tools
cannot be produced---but it eliminates the glue-code failure mode
entirely. Whether hybrid designs can achieve a better trade-off is an
open question; the answer likely depends on how reliably language
models can assemble correct pipelines around trusted library calls,
and on the target user's ability to review what was assembled.

\begin{table}[htbp]
\centering
\begin{threeparttable}
\caption{Comparison with code-generation approaches}
\label{tab:system-comparison}
\small
\begin{tabular}{@{}lp{2.5cm}p{2.4cm}p{2.4cm}p{2.4cm}@{}}
\toprule
 & \pkg{p2a} & Econ.\ AI & Data Interpr. & LAMBDA \\
\midrule
LLM role & Tool selection & Enhanced code gen & Code gen + exec & Agent orchestration \\
Computation & Pre-validated & LLM-generated & LLM-generated & LLM-generated \\
Local execution & Yes (Ollama) & No & No & No \\
Validated & Yes (vs.\ R) & No & No & No \\
Method flexibility & Fixed library (${\sim}200$ methods) & Any \proglang{R} method & Any \proglang{Python} method & Any \proglang{Python} method \\
Extensibility & Requires Rust impl. & Extends with model & Extends with model & Extends with model \\
\bottomrule
\end{tabular}
\begin{tablenotes}[flushleft]
\item \emph{Note:} Econ.\ AI = Econometrics AI
Agent~\citep{Chen+etal:2025}; Data Interpr.\ = Data
Interpreter~\citep{Hong+etal:2024}; LAMBDA~\citep{Sun+etal:2025:LAMBDA}.
The trade-off: code-generation systems offer unlimited method
flexibility at the cost of unvalidated computation; \pkg{p2a}
constrains coverage to ensure correctness.
\end{tablenotes}
\end{threeparttable}
\end{table}

Three limitations of the approach are fundamental, not fixable by
better engineering. First, statistical expertise is still required:
chat-first analytics assists with method selection and execution, not
with research design or causal identification. A user who requests
``estimate the effect of X on Y'' receives a regression, but the
system does not validate whether a causal interpretation is warranted.
LLM explanations may echo causal language from the user's prompt even
when the analysis provides only associational evidence. Users remain
responsible for ensuring that their identification strategy supports
their claims.

Second, as in any conversational interface, method selection is
imperfect. Tool selection works well on curated prompts
(Section~\ref{sec:evaluation}), but users should verify that the
system understood their request correctly---particularly for ambiguous
specifications where multiple methods could apply.

Third, pre-validation trades coverage for correctness. When a user
requests a method outside the implemented set---Bayesian inference,
structural equation modeling, or a custom estimator---the LLM may
silently map the request to the closest available alternative rather
than reporting unavailability---the symmetric failure mode to
code generation's silent errors: instead of running wrong code, the
system runs the wrong method. The risk is mitigated by the structured
tool schema (which makes the selected method visible to the user) and
by the LLM's ability to explain substitutions, but users must still
verify that the method matches their intent.\footnote{This limitation
will diminish as the library grows. The current implementation covers
over 200 methods after roughly one year of development; extending
coverage to additional methods is straightforward within the existing
architecture.}

These limitations mean that chat-first analytics is a useful tool,
not an autonomous analyst. Users should verify tool selection, check
extracted parameters, and apply domain knowledge to interpret
results---the same discipline required when using any statistical
software, with the added need to verify that the system understood the
request correctly.
