%% =============================================================================
%% EXAMPLES
%% =============================================================================

\section{Examples} \label{sec:examples}

This section demonstrates \pkg{prompt2analytics} through
two interfaces: the natural language chat interface powered by large language
models, and the command-line interface for scripted workflows. Both interfaces
access the same underlying \proglang{Rust} analytics engine.

We use the Grunfeld investment data~\citep{Baltagi:2013}, a classic
panel dataset with 200 observations (10 firms over 20 years) containing
firm investment, market value, and capital stock. The dataset is included
in the \pkg{plm} package; we provide it as \code{grunfeld.csv} in the
supplementary materials.

\noindent The early examples show the system working as intended; later
examples (Section~\ref{sec:examples-realistic}) demonstrate failure
modes---incorrect tool selection, missing parameters, and ambiguous
prompts---to set realistic expectations before the formal evaluation in
Section~\ref{sec:evaluation}.  Appendix~\ref{app:extended-examples}
provides additional examples including data
cleaning workflows and a comparison with equivalent \proglang{R} code.

%% -----------------------------------------------------------------------------
\subsection{Chat interface} \label{sec:examples-chat}

The \code{p2a chat} command provides a terminal REPL for conversational
analytics.  It connects to a running \pkg{p2a-mcp} HTTP server and streams
responses via server-sent events, so all LLM calls, tool execution, and
session management happen server-side.  Users describe their analysis goals
in plain language, and the LLM translates these into structured tool calls
executed by the MCP server.  Tool invocations are shown inline as they
execute (tool name, arguments, elapsed time), followed by the LLM's
narrative response.

The transcript below was recorded from an actual terminal session using
Claude Sonnet~4.6.\footnote{The session was generated non-interactively
via \code{p2a chat --script prompts.txt --record transcript.txt --provider
anthropic --model claude-sonnet-4-6}.  The \code{--script} flag reads
prompts from a file; \code{--record} saves a plain-text transcript.  The
full transcript and prompt file are included in the supplementary materials.
Other models (GPT-4.1~Mini, Ministral~3B, etc.)\ produce similar results
with minor variations in explanation style.}
The outputs shown here are lightly condensed for space---the full,
unedited transcript is available in \code{paper/transcripts/}.

A typical session begins by loading data and examining its structure:

\begin{Sinput}
p2a> Load the Grunfeld investment dataset from grunfeld.csv and show me
     what's in it.
\end{Sinput}

\begin{Soutput}
  [tool] load_dataset (name="grunfeld", path="grunfeld.csv")  done 6ms
  [tool] head_dataset (dataset="grunfeld", n=10)              done 1ms
  [tool] describe_dataset (dataset="grunfeld")                done 0ms

Dataset Overview: 200 rows, 5 columns, no missing values.

| Column  | Type    | Description                                  |
|---------|---------|----------------------------------------------|
| firm    | Integer | Firm identifier (1-10, 10 firms)             |
| year    | Integer | Year (1935-1954)                             |
| inv     | Float   | Gross investment (dependent variable, M$)    |
| value   | Float   | Market value of the firm (M$)                |
| capital | Float   | Capital stock (net plant & equipment, M$)    |

Classic panel dataset: 10 firms x 20 years = 200 balanced observations.
\end{Soutput}

The next example shows cross-sectional regression with heteroskedasticity-consistent standard errors.%
\footnote{HC1 is the default robust standard error estimator when robust SEs are
requested. HC1 applies a degrees-of-freedom correction $n/(n-k)$ to HC0,
improving small-sample performance.}

\begin{Sinput}
p2a> Run a regression of investment on market value and capital stock.
     Use robust standard errors.
\end{Sinput}

\begin{Soutput}
  [tool] regression_ols (dataset="grunfeld", y="inv",
           x=["value","capital"])                              done 5ms

OLS Regression Results (HC1 robust standard errors)
Dependent variable: inv

| Variable    |    Coef |  Std Err |      t | P>|t| |
|-------------|--------:|---------:|-------:|------:|
| (Intercept) | -42.714 |   11.575 | -3.690 | 0.000 |
| value       |   0.116 |    0.007 | 16.967 | 0.000 |
| capital     |   0.231 |    0.049 |  4.721 | 0.000 |

R-squared: 0.812,  Adj. R-squared: 0.810
F-statistic: 426.58 (p < 0.001),  Observations: 200

Both market value and capital stock are highly significant predictors of
investment.  A $1M increase in market value is associated with $0.116M
more in gross investment, holding capital constant.
This is pooled OLS---it ignores the panel structure (firm fixed effects).
\end{Soutput}

For panel data, users can request fixed effects estimation in natural language:

\begin{Sinput}
p2a> The data has a panel structure. Run a fixed effects regression
     controlling for firm-specific effects.
\end{Sinput}

\begin{Soutput}
  [tool] panel_fixed_effects (dataset="grunfeld", y="inv",
           x=["value","capital"], entity_var="firm")           done 2ms

Fixed Effects (Within) Estimation
Dependent variable: inv,  Entity fixed effects: firm (10 levels)

| Variable |    Coef |  Std Err |      t | P>|t| |
|----------|--------:|---------:|-------:|------:|
| value    |   0.110 |    0.012 |  9.288 | 0.000 |
| capital  |   0.310 |    0.017 | 17.867 | 0.000 |

R-squared (within): 0.767,  Observations: 200,  Groups: 10
F-statistic: 309.01 (p < 0.001)

The capital stock coefficient increases from 0.231 to 0.310 after
controlling for firm fixed effects---pooled OLS was underestimating the
within-firm effect.  Market value is slightly lower (0.116 -> 0.110),
suggesting some of the pooled effect was driven by cross-firm differences.
\end{Soutput}

Users can request specification tests without knowing the technical details:

\begin{Sinput}
p2a> Should I use fixed effects or random effects for this analysis?
\end{Sinput}

\begin{Soutput}
  [tool] hausman_test (dataset="grunfeld", y="inv",
           x=["value","capital"], entity_var="firm")           done 60ms

Hausman Test Results
H0: Random effects is consistent and efficient
H1: Fixed effects is consistent; random effects is inconsistent

Chi-squared statistic: 0.618
Degrees of freedom: 2
P-value: 0.734

Coefficient comparison:
| Variable | Fixed Effects | Random Effects | Difference |
|----------|-------------:|---------------:|-----------:|
| value    |        0.110 |          0.110 |      0.000 |
| capital  |        0.310 |          0.308 |      0.002 |

With p = 0.734, we fail to reject H0---random effects appears
consistent. However, the Hausman test has limited power with only
10 entities. Economically, unobserved firm characteristics likely
correlate with market value and capital, arguing for fixed effects.
Since FE and RE estimates are nearly identical here, the practical
choice matters less for this dataset.
\end{Soutput}

\noindent In this example, the test statistics ($\chi^2$, $p$-value,
degrees of freedom) come from the \proglang{Rust} backend; the interpretive
caveat about limited power with few entities is generated by the LLM.
This division is typical: the computation layer returns structured results
and simple diagnostics, and the language model adds contextual
interpretation---including caveats about assumptions and limitations that
depend on the specific results observed.

For datasets with multiple categorical dimensions, users can request
multi-way fixed effects:

\begin{Sinput}
p2a> Control for both firm and year fixed effects in the regression.
\end{Sinput}

\begin{Soutput}
  [tool] panel_hdfe (dataset="grunfeld", y="inv",
           x=["value","capital"], fe=["firm","year"])          done 5ms

High-Dimensional Fixed Effects Estimation
Dependent variable: inv
Fixed effects: firm (10 levels), year (20 levels)

| Variable |    Coef |  Std Err |      t | P>|t| |
|----------|--------:|---------:|-------:|------:|
| value    |   0.118 |    0.019 |  6.137 | 0.000 |
| capital  |   0.358 |    0.054 |  6.579 | 0.000 |

R-squared (within): 0.720,  Observations: 200
F-statistic: 217.44 (p < 0.001),  Convergence: 2 iterations

Adding year fixed effects absorbs common macro shocks (recessions, WWII,
post-war boom), revealing a stronger capital-investment relationship
(0.358, up from 0.310 in the firm-FE-only model).
\end{Soutput}

Users can also request diagnostic tests in natural language:

\begin{Sinput}
p2a> Check the regression assumptions for the OLS model.
\end{Sinput}

\begin{Soutput}
  [tool] regression_diagnostics (dataset="grunfeld", y="inv",
           x=["value","capital"])                              done 22ms

Jarque-Bera Test (Normality of residuals)
Statistic: 89.57,  P-value: 0.000
Conclusion: Residuals deviate significantly from normality.

Breusch-Pagan Test (Heteroskedasticity)
LM statistic: 55.25,  P-value: 0.000
Conclusion: Evidence of heteroskedasticity; robust SEs recommended.

Durbin-Watson Statistic (Autocorrelation)
Statistic: 0.358
Interpretation: Strong positive serial autocorrelation (expected in
panel data where investment decisions persist over time).

Variance Inflation Factors
value: 1.31,  capital: 1.31
No multicollinearity detected (VIF < 5).
The high condition number (2,458) is driven by variable scale, not
true collinearity.
\end{Soutput}


%% -----------------------------------------------------------------------------
\subsection{Command-line interface} \label{sec:examples-cli}

The command-line interface provides programmatic access for scripted
workflows.\footnote{The output shown in this section was captured from
actual CLI invocations; the full transcript is in
\code{paper/transcripts/section4\_2\_cli\_output.txt}.}
Commands follow a hierarchical structure with sessions maintaining state
across invocations:

\begin{Sinput}
$ p2a --session analysis.json data load grunfeld.csv --name grunfeld
\end{Sinput}

\begin{Soutput}
Dataset: grunfeld
  Rows: 200
  Columns: 5
  Column names: firm, year, inv, value, capital
\end{Soutput}

\begin{Sinput}
$ p2a --session analysis.json reg ols grunfeld \
    -y inv -x value capital --robust hc1
\end{Sinput}

\begin{Soutput}
OLS Regression Results
============================================================
Variable         Coefficient   Std. Error    t-value    p-value
------------------------------------------------------------
(Intercept)       -42.714369    11.574701     -3.690     0.0003***
value               0.115562     0.006811     16.967     0.0000***
capital             0.230678     0.048866      4.721     0.0000***
------------------------------------------------------------
R-squared: 0.812408
Adj. R-squared: 0.809537
Observations: 200
\end{Soutput}

\begin{Sinput}
$ p2a --session analysis.json panel fe grunfeld \
    -y inv -x value capital --entity firm
\end{Sinput}

\begin{Soutput}
Fixed Effects (Within) Estimation Results
============================================================
Variable         Coefficient   Std. Error    t-value    p-value
------------------------------------------------------------
value               0.110124     0.011857      9.288     0.0000***
capital             0.310065     0.017355     17.867     0.0000***
------------------------------------------------------------
R-squared: 0.766758,  Observations: 200,  Groups: 10
F-statistic: 309.0142 (p-value: 0.0000)
\end{Soutput}

All commands support JSON output (\code{-F json}) for integration with
other tools and scripts:

\begin{Sinput}
$ p2a -F json reg ols grunfeld -y inv -x value capital --robust hc1
\end{Sinput}

\begin{Soutput}
{
  "method": "OLS Regression",
  "r_squared": 0.8124,
  "adj_r_squared": 0.8095,
  "n_observations": 200,
  "coefficients": [
    {"name": "(Intercept)", "coefficient": -42.714,
     "std_error": 11.575, "t_value": -3.690, "p_value": 0.0003},
    {"name": "value", "coefficient": 0.1156,
     "std_error": 0.0068, "t_value": 16.967, "p_value": 0.0},
    {"name": "capital", "coefficient": 0.2307,
     "std_error": 0.0489, "t_value": 4.721, "p_value": 0.0}
  ]
}
\end{Soutput}

\noindent This enables integration with analysis pipelines, for example,
piping results to \code{jq} for extraction or to \proglang{Python}/\proglang{R}
scripts for further processing. The session mechanism records all commands
and can export them as an executable shell script for exact reproduction.

%% -----------------------------------------------------------------------------
\subsection{Realistic use and error handling} \label{sec:examples-realistic}

The preceding examples show successful interactions. Real use involves
more iteration. This section demonstrates a multi-step
difference-in-differences workflow---including method selection,
clustered standard errors, and diagnostic issues---using a synthetic
employee training dataset.\footnote{The dataset contains 1,200
observations (200 employees $\times$ 6 periods) with treatment/control
groups, perfectly correlated covariates (for multicollinearity), and a
department with zero bonus rate (for separation). The transcript was
recorded via \code{p2a chat --script prompts.txt --record transcript.txt}.}

Given a description of the research design, the LLM correctly selects
difference-in-differences and proactively clusters standard errors at the
employee level---recognizing this as standard practice for panel DiD:

\begin{Sinput}
p2a> Analyze whether the training program affected employee performance.
     The data has employees observed before and after training, with a
     control group that didn't receive training.
\end{Sinput}

\begin{Soutput}
  [tool] diff_in_diff (dataset="training", dep_var="performance",
           post_var="post_training", treatment_var="training_group",
           cluster_var="employee_id")                                done 6ms
  [tool] panel_fixed_effects (dataset="training",
           entity_var="employee_id", y="performance",
           x=["post_training","training_group"])                     done 5ms

Difference-in-Differences Estimation (clustered by employee_id)

| Group      | Pre-Training | Post-Training | Change |
|------------|-------------:|--------------:|-------:|
| Control    |        3.002 |         3.127 | +0.125 |
| Treated    |        3.424 |         3.763 | +0.339 |
| DiD Effect |              |               | +0.213 |

The ATT estimate is +0.213 (SE: 0.035, t = 6.06, p < 0.001),
with standard errors clustered across 200 employees. The training
program improved performance by approximately 0.21 points---roughly
a 6\% lift relative to the treated group's pre-training mean.
\end{Soutput}

\noindent The LLM passed \code{cluster\_var="employee\_id"} without
being asked---it recognized that panel DiD requires clustered inference.
The user can verify this by requesting explicit clustering:

\begin{Sinput}
p2a> Run the difference-in-differences again but with standard errors
     clustered at the employee level.
\end{Sinput}

\begin{Soutput}
  [tool] diff_in_diff (dataset="training", dep_var="performance",
           post_var="post_training", treatment_var="training_group",
           cluster_var="employee_id")                                done 9ms

| Variable              |    Coef | Cluster SE |      t | P>|t| |
|-----------------------|--------:|-----------:|-------:|------:|
| training_group        |   0.422 |      0.072 |  5.87  | 0.000 |
| post_training         |   0.125 |      0.023 |  5.37  | 0.000 |
| DiD (ATT)             |   0.213 |      0.035 |  6.06  | 0.000 |

The ATT remains +0.213 with clustered SE of 0.035 (p < 0.001).
\end{Soutput}

When perfectly correlated covariates are added, the system detects
multicollinearity and explains both the symptoms and the fix:

\begin{Sinput}
p2a> Add years_since_hire to the model alongside tenure.
\end{Sinput}

\begin{Soutput}
  [tool] regression_clustered (cluster1="employee_id",
           dataset="training", y="performance",
           x=["training_group","post_training","did",
               "tenure","years_since_hire"])                         done 14ms

Multicollinearity Warning: tenure and years_since_hire are
near-perfectly collinear

| Variable         | Estimate | Std. Error |
|------------------|---------:|-----------:|
| tenure           |    0.438 |        --- |
| years_since_hire |   -0.438 |        --- |
| DiD (ATT)        |    0.213 |      0.035 |

The two variables are essentially measuring the same quantity
(years_since_hire = tenure + constant). Standard errors cannot be
computed individually. The DiD estimate remains stable at +0.213
(p < 0.001)---the treatment effect is unaffected.

Recommendation: include only one of the two variables.
\end{Soutput}

\noindent The system diagnoses the problem from the numerical symptoms
rather than simply reporting a matrix singularity error. The LLM's
explanation distinguishes between the unreliable individual coefficients
and the stable treatment effect---a nuance that requires understanding
both the computation and the research design.

The session continues with a logit model for bonus receipt with
department fixed effects---a discrete choice model with high-dimensional
FE estimated via IRLS:

\begin{Sinput}
p2a> Fit a logit model for bonus with department fixed effects.
\end{Sinput}

\begin{Soutput}
  [tool] feglm (dataset="training", family="logit",
           fe=["department"],
           x=["performance","tenure","training_group",
               "post_training"])                                     done 102ms

Logit Model for Bonus (Department FE, 5 levels absorbed)
Observations: 1,200 (436 bonuses),  Pseudo R-squared: 0.082

| Variable       |    Coef | Std Err |      z | P>|z| |
|----------------|--------:|--------:|-------:|------:|
| performance    |   0.305 |   0.105 |  2.90  | 0.004 |
| tenure         |   0.000 |   0.011 |  0.03  | 0.977 |
| training_group |  -0.004 |   0.138 | -0.03  | 0.978 |
| post_training  |   0.118 |   0.128 |  0.92  | 0.359 |

Performance is the only significant predictor of bonus receipt.
The training program does not independently affect bonuses beyond
its indirect effect through improved performance.
\end{Soutput}

\noindent This example spans three method categories in a single
conversation: DiD (causal inference), OLS with diagnostics (regression),
and FEGLM (discrete choice)---demonstrating how the conversational
interface enables fluid transitions between methods as the analysis
evolves.

%% -----------------------------------------------------------------------------
\subsection{Beyond regression: time series} \label{sec:examples-ts}

The system covers domains beyond regression. A time series workflow
using quarterly GDP growth data (1954--2023, 280 observations)
illustrates method progression within a single conversation:
stationarity testing, univariate modeling, then residual diagnostics.

\begin{Sinput}
p2a> Load gdp_quarterly.csv as gdp and check whether GDP growth
     is stationary.
\end{Sinput}

\begin{Soutput}
  [tool] load_dataset (name="gdp",
           path="validation/datasets/gdp_quarterly.csv")            done 49ms
  [tool] timeseries_pp_test (column="gdp_growth", dataset="gdp")    done 24ms

Phillips-Perron Test (gdp_growth)
Z(tau) statistic: -211.82,  P-value: < 0.0001
GDP growth is stationary. No differencing needed.
\end{Soutput}

\begin{Sinput}
p2a> Fit an ARIMA model to GDP growth. Let the system pick the order.
\end{Sinput}

\begin{Soutput}
  [tool] ts_arima_fit (column="gdp_growth", d=0, dataset="gdp",
           p=1, q=1)                                                done 54ms

ARIMA(1,0,1): gdp_growth_t = 0.233 + 0.549 * gdp_growth_{t-1}
              + e_t - 0.249 * e_{t-1}
AIC: 657.92,  Observations: 280
\end{Soutput}

\begin{Sinput}
p2a> Check for residual autocorrelation in the ARIMA model.
\end{Sinput}

\begin{Soutput}
  [tool] regression_bgtest (dataset="gdp", order=4,
           x=["year","quarter"], y="gdp_growth")                    done 7ms
  [tool] ts_arima_fit (column="gdp_growth", d=0, dataset="gdp",
           p=2, q=2)                                                done 11ms

Breusch-Godfrey Test (up to lag 4)
LM statistic: 32.12,  p-value: < 0.0001
Residual autocorrelation detected.

Remediation: ARIMA(2,0,2) was fitted but does not improve AIC
(660.6 vs 657.9). The remaining autocorrelation is statistically
significant but economically modest (auxiliary R-squared = 0.115).
Recommendation: retain ARIMA(1,0,1) and apply HAC standard errors
for downstream inference.
\end{Soutput}

\noindent Order selection is delegated to the LLM, which chooses
$p$, $d$, $q$ based on the stationarity test result and its training
knowledge. This is not equivalent to \proglang{R}'s \code{auto.arima()},
which performs systematic grid
search~\citep{Hyndman+Khandakar:2008}. The LLM autonomously checks a
higher-order specification and reports that the parsimony--fit tradeoff
favors the simpler model.

Appendix~\ref{app:extended-examples} provides additional examples
including data cleaning workflows and a side-by-side comparison with
equivalent \proglang{R} code.

These examples illustrate what the system can do; the next section
measures how reliably it does so. We evaluate tool selection accuracy
across six models and numerical correctness of the underlying
\proglang{Rust} implementations.
