%% =============================================================================
%% INTRODUCTION
%% =============================================================================

\section{Introduction} \label{sec:intro}

The release of ChatGPT in late 2022 changed how many researchers and analysts
approach data analysis. Rather than consulting documentation, writing code from
scratch, or learning domain-specific syntax, users began asking questions in
plain English: ``run a regression of wages on education and experience,'' ``test
whether these two groups have different means,'' or ``fit an ARIMA model to this
time series.'' Natural language interfaces to statistical software have
been explored for decades, from early database query
systems~\citep{Androutsopoulos+etal:1995} to commercial tools like SPSS and SAS. Large
language models (LLMs) are more flexible and accessible than these predecessors,
but they also raise concerns about accuracy, reproducibility, and data privacy.

This paper describes \emph{chat-first data analytics}: an application of
tool-augmented language model
techniques~\citep{Schick+etal:2023, Yao+etal:2023} to statistical and econometric analysis.
Natural language conversation is the primary interface
to statistical methods, while pre-implemented algorithms perform the
actual computation. Language models perform well at understanding intent and
selecting appropriate methods, but they should not be trusted to generate
correct statistical code. Separating interpretation (what does the user want?)
from computation (how do we compute it correctly?) allows chat-first analytics
to provide accessible conversational interfaces while maintaining the reliability
that scientific work requires.

\subsection{The current landscape}

The use of LLMs for data analysis has grown
rapidly~\citep{Nejjar+etal:2024, Sun+etal:2025:Survey, Korinek:2025}. The appeal is
clear: natural language eliminates the need to learn programming syntax, allowing
domain experts to describe analytical goals directly.

The pitfalls, however, are substantial. LLM-generated code may run without error
but compute the wrong quantity---requesting ``clustered standard errors'' might
yield robust errors instead, or a panel model might omit fixed effects.
\cite{Ludwig+Mullainathan+Rambachan:2025} emphasize the risks of ``training
leakage'' and the difficulty of validating generated code. Users who lack the
expertise to write code often lack the expertise to audit it.

Two additional concerns dominate: \emph{data privacy}---sensitive information
flows to third-party servers with unclear policies---and \emph{reproducibility}---
conversational sessions are ephemeral, with the same prompt yielding different
code as models update.

These concerns are not confined to research settings. Enterprise analytics vendors
have similarly retreated from ``chat with your data'' approaches after encountering
systematic failures: inconsistent metric definitions across data sources, hidden
business logic inaccessible to language models, and outputs that fail basic audit
requirements. When identical questions yield different answers depending on which
dataset is queried, or when results cannot be traced back to underlying
calculations, user trust erodes rapidly. The pattern suggests that free-form
conversational interfaces to data---whether for business intelligence or
statistical analysis---require architectural constraints that naive
implementations lack.

\subsection{The opportunity}

Despite these concerns, the underlying demand is legitimate and likely to
persist. Researchers want to focus on their substantive questions (Does
education affect wages? Did the policy reduce crime? Is this treatment
effective?) rather than on implementation details. The cognitive burden of
translating analytical intent into software-specific syntax is real, and
reducing this burden frees mental resources for the questions
that actually matter.

Natural language interfaces may also enable exploratory dialogue that code-based
workflows discourage. A researcher might begin with
``show me the relationship between X and Y,'' examine the scatterplot, then ask
``control for Z,'' inspect the regression output, and continue ``are there
outliers driving this?'' Each step builds on the previous, with the language
model maintaining context across the conversation. This mirrors how experienced
analysts actually think about data. Whether this translates to better user
experience remains to be tested in user studies.

A further opportunity comes from reconsidering the computational architecture.
When ChatGPT or Claude perform data analysis, they generate \proglang{Python}
code that executes in ephemeral virtual machines with limited computational
resources. This reflects the code-generation approach: the LLM writes a program,
and a sandboxed environment runs it. If we instead route requests through a
structured protocol like MCP, the computational backend need not be constrained
by what an LLM can generate or what a sandboxed \proglang{Python} environment
can efficiently execute. We can implement statistical methods in compiled
languages optimized for numerical performance. \proglang{Rust} offers memory
safety guarantees enforced at compile time alongside performance comparable to
\proglang{C} and \proglang{Fortran}, the traditional languages of scientific
computing. MCP thus enables high-performance analytics that are difficult to
achieve with code generation.

The opportunity is to design systems that preserve the accessibility and
interactivity of conversational interfaces while addressing concerns about
correctness, privacy, reproducibility, and performance. Recent work has begun
exploring this direction: \cite{Chen+Shen+Ma:2025} develop an ``Econometrics AI
Agent'' that augments LLMs with domain knowledge to improve completion rates on
econometric tasks, while \cite{Hong+etal:2024} present a general-purpose ``Data
Interpreter'' agent. Our approach differs by emphasizing pre-validated
implementations and local execution rather than enhanced code generation.

\subsection{Chat-first data analytics}

We use \emph{chat-first data analytics} to describe an approach that applies
tool-augmented language model techniques~\citep{Schick+etal:2023, Yao+etal:2023}
to statistical and econometric analysis. This extends the natural language
interface pattern---developed for database queries, data visualization, and
general-purpose AI assistants---to a new domain, applying established methods
rather than introducing novel techniques. The approach has two defining
properties and one important design choice:

\begin{enumerate}
  \item \textbf{Natural language as the primary interface.} Users describe
    their analytical goals in plain language rather than programming syntax.
    The system interprets requests, selects appropriate methods, and explains
    results conversationally.
  \item \textbf{Pre-implemented, verified statistical methods.} The language
    model does not generate statistical code. Instead, it invokes pre-built,
    tested implementations of standard methods. Correctness is established
    once, through validation against reference implementations, rather than
    verified anew for each user request.
  \item \textbf{Local data processing} (design choice). In our implementation,
    bulk data never leaves the user's machine; the language model receives
    only metadata and aggregated results. This is not definitionally required
    for chat-first analytics---one could build such a system with cloud-based
    computation---but we consider it essential for addressing the privacy
    concerns that motivate much of this work.
\end{enumerate}

This architecture addresses each concern identified above. Correctness is
ensured by using validated implementations rather than generated code.
Privacy is protected by keeping data local. Reproducibility is achieved
because the same method invocation produces identical results---the
stochastic element of the language model affects only interpretation, not
computation.

Chat-first differs from both traditional statistical software
and from naive LLM-assisted analysis. Unlike \proglang{R}, \proglang{Stata},
or \proglang{Python}, chat-first systems do not require users to learn
domain-specific syntax. Unlike asking ChatGPT to write code, chat-first
systems do not trust the language model with numerical computation.
The language model is a translator, converting natural language
intent into structured method calls, while verified software performs
the actual analysis. This approach builds on tool-augmented
language models~\citep{Schick+etal:2023, Qin+etal:2024, Liu+etal:2025:ToolACE},
which can reliably select and invoke external tools when provided with
appropriate schemas~\citep{Qu+etal:2025}.

\subsection{This paper}

This paper presents \pkg{prompt2analytics}, an implementation of
chat-first data analytics. The software includes an econometrics
library in pure \proglang{Rust}, exposed through the Model Context
Protocol (MCP) for integration with language models. Users interact through
natural language; the language model interprets requests and generates
structured tool calls; the \proglang{Rust} engine executes validated
statistical methods; and results flow back through the language model for
explanation.

The implementation provides methods across the core areas of applied
econometrics: regression analysis with heteroskedasticity-consistent and
clustered standard errors, panel data models including high-dimensional
fixed effects, instrumental variables estimation, difference-in-differences,
discrete choice models, time series analysis, and standard hypothesis tests.
All implementations are validated against established \proglang{R} packages
to ensure numerical accuracy.

The remainder of the paper is organized as follows. Section~\ref{sec:design}
elaborates the design principles and includes practical accuracy expectations
(Section~\ref{sec:accuracy-expectations}), which readers should consult before
interpreting the polished examples that follow. Section~\ref{sec:architecture}
describes the software architecture of \pkg{prompt2analytics}, including the
core library, implemented methods, user interfaces, and communication flow.
Section~\ref{sec:examples} demonstrates the chat-first workflow through worked
examples, including realistic failure cases. Section~\ref{sec:evaluation}
evaluates LLM tool selection accuracy and numerical validation against
\proglang{R}.
Section~\ref{sec:comparison} presents validation results comparing
\pkg{prompt2analytics} to established \proglang{R} packages, and
Section~\ref{sec:performance} provides detailed performance benchmarks.
Section~\ref{sec:local} discusses fully local deployment for
privacy-sensitive applications. Section~\ref{sec:conclusion} concludes with
limitations and future directions.

Table~\ref{tab:jss-mapping} maps this structure to JSS's expected sections.

\begin{table}[htbp]
\centering
\caption{Mapping of paper sections to JSS expected structure}
\label{tab:jss-mapping}
\small
\begin{tabular}{llc}
\toprule
JSS Expected Section & Paper Section & Section \\
\midrule
Introduction / Motivation & Introduction & \ref{sec:intro} \\
Background / Related work & Background and design principles & \ref{sec:design} \\
Software design & Software architecture & \ref{sec:architecture} \\
Illustrations / Examples & Examples & \ref{sec:examples} \\
Computational details & Evaluation + Comparison & \ref{sec:evaluation}, \ref{sec:comparison} \\
Performance & Performance benchmarks & \ref{sec:performance} \\
--- & Local deployment and privacy & \ref{sec:local} \\
Summary / Outlook & Conclusion & \ref{sec:conclusion} \\
\bottomrule
\end{tabular}
\end{table}

\noindent Section~\ref{sec:local} (Local deployment) has no direct JSS analog
but addresses data privacy, a central design motivation, and provides
deployment guidance specific to local LLM configurations.

\subsection{The statistical software landscape}

Econometrics software has long been dominated by
\proglang{R}~\citep{R}, \proglang{Stata}, and increasingly \proglang{Python}.
\proglang{R} has over 200 econometrics-related
packages~\citep{Michelaki+Tsagris:2025}. Numerical accuracy remains a concern
across platforms~\citep{McCullough:1997, McCullough+Vinod:1999}, motivating
careful validation.

\proglang{Rust} has become a reasonable platform for scientific computing,
combining compiled-language performance with memory safety guarantees enforced
at compile time~\citep{Rust}. The ecosystem has matured, with
\pkg{polars} for DataFrames~\citep{polars} and \pkg{faer} for linear algebra.
Until now, no econometrics library has existed for \proglang{Rust}.
\pkg{prompt2analytics} addresses this gap.

\subsection{Software availability}

\pkg{prompt2analytics} is available as open-source software under the MIT
license, which permits commercial use, modification, and redistribution with
attribution. The source code, pre-built binaries, installation instructions,
and documentation are available at the project repository. Contributions are
welcome via pull requests; the repository includes contribution guidelines
and a code of conduct.

\begin{center}
\url{https://github.com/umatter/prompt2analytics}
\end{center}

\noindent Supplementary materials accompanying this paper include:
\begin{itemize}
  \item The 87 test cases used for LLM tool selection evaluation
    (Section~\ref{sec:evaluation}), provided as JSON files with prompts,
    expected tools, and acceptable alternatives
  \item Benchmark scripts for both \proglang{R} and \proglang{Rust}
    implementations, enabling reproduction of performance comparisons
    (Section~\ref{sec:performance})
  \item Validation test suites comparing numerical output against
    \proglang{R} reference implementations (Section~\ref{sec:comparison})
\end{itemize}

\noindent These materials are located in the repository's \code{paper/}
directory and can be executed using the provided Makefile.
Appendix~\ref{app:reproducibility} provides detailed reproduction instructions,
version information, and containerized build procedures.

