%% =============================================================================
%% DESIGN PHILOSOPHY
%% =============================================================================

\section{Design philosophy} \label{sec:design}

Traditional statistical software requires users to learn domain-specific syntax:
\proglang{R} formulas, \proglang{Stata} commands, or \proglang{Python} APIs.
Users must translate their analytical intent into software-specific instructions.
\pkg{prompt2analytics} inverts this: users describe their goals in natural
language, and the system determines the appropriate statistical method.

\subsection{Chat-first analytics}

The design principle is \emph{chat-first analytics}:
the primary interface is conversational rather than programmatic. A user
might request ``estimate the effect of education on wages, controlling for
experience and industry, with robust standard errors'' without specifying
whether this requires OLS, fixed effects, or instrumental variables. The
system interprets the request, selects appropriate methods, executes the
analysis, and returns results with natural language explanation.

This approach builds on a long history of natural language interfaces for
data. Early systems like LUNAR \citep{Woods:1973} demonstrated that users
could query databases in English, though these systems were limited to
specific domains and required careful phrasing
\citep{Androutsopoulos+etal:1995}. The text-to-SQL paradigm, exemplified
by benchmarks like Spider \citep{Yu+etal:2018}, has driven substantial
progress in translating natural language to structured queries
\citep{Li+etal:2024}. Similarly, systems like NL4DV \citep{Narechania+etal:2021}
enable natural language specification of data visualizations.

Our approach differs from these predecessors in two key respects. First,
we target statistical analysis rather than data retrieval or visualization
alone---the system must select among econometric methods, not just translate
queries to SQL. Second, we leverage large language models as the interpretive
layer, providing flexibility that rule-based systems cannot match.

This conversational approach offers several advantages. It lowers the barrier
to entry for users unfamiliar with econometric software conventions. It
allows domain experts to focus on their research questions rather than
implementation details. And it enables exploratory analysis through
dialogue: users can refine their analysis iteratively based on results
and suggestions from the system.

\subsection{Positioning this work}

\pkg{prompt2analytics} builds on a long tradition of interface innovation in
statistical software, from the Wilkinson-Rogers formula notation
\citep{Wilkinson+Rogers:1973} to text-to-SQL systems \citep{Yu+etal:2018} and
AutoML pipelines \citep{Feurer+etal:2015}. Rather than inventing new notation,
it accepts natural language descriptions of analytical intent, trading the
precision of formal DSLs for accessibility.

The contribution is software engineering rather than methodological
innovation: (1) providing a \proglang{Rust} econometrics library,
(2) exposing implementations through a standardized LLM integration protocol,
and (3) demonstrating that chat-first interfaces to validated backends can
address correctness and privacy concerns. The implementations follow established
algorithms validated against \proglang{R} reference packages.
Appendix~\ref{app:related-work} discusses related work on statistical interfaces.

\subsection{LLM limitations in chat-first systems}

Chat-first does not replace statistical expertise. Users
must still understand their data, formulate sensible research questions,
and interpret results critically. The system assists with method selection
and execution, not with research design or causal identification.

A related concern is the handling of causal language. Users may request
analyses using causal terminology (``estimate the effect of X on Y'')
without satisfying identification assumptions. The system does not
validate whether causal claims are supported by the research design;
it executes the requested statistical procedure and returns results.
LLM explanations may use causal language that echoes user prompts even
when the underlying analysis provides only associational evidence.
Users bear responsibility for ensuring that causal interpretations are
warranted by their identification strategy, not merely by the statistical
method employed.

\subsubsection{Identification warnings}

The system implements a diagnostics module that generates data-driven
warnings when estimation results suggest potential violations of
identifying assumptions. These warnings are non-blocking---results are
always returned to the user---but they highlight conditions that warrant
careful interpretation.

For \emph{instrumental variables}, the system checks first-stage
$F$-statistics against the Stock-Yogo threshold of 10
\citep{Stock+Yogo:2005}. Values below 10 trigger weak instrument warnings
with severity scaled by the degree of weakness ($F < 5$ generates critical
warnings). When models are overidentified, the Sargan test is automatically
computed and rejection ($p < 0.05$) triggers exclusion restriction warnings.

For \emph{difference-in-differences}, staggered designs include a pre-trend
test (joint test that pre-treatment ATT estimates are zero). Rejection
generates parallel trends warnings. The system also checks for significant
individual pre-treatment effects that may indicate anticipation.

For \emph{matching and IPW}, balance diagnostics check whether post-matching
standardized mean differences exceed 0.1, the conventional threshold for
acceptable balance \citep{Austin:2011}. Propensity score distributions are
examined for positivity violations---scores near 0 or 1 imply extreme
weights. Match rates are reported to flag limited common support.

For \emph{regression discontinuity}, effective sample sizes within the
bandwidth are checked against minimum thresholds ($n < 50$ per side triggers
warnings). Bandwidth sensitivity is flagged to encourage robustness checks
at alternative bandwidths.

Each warning includes: (1) a severity level (info, caution, warning, critical),
(2) a diagnostic message explaining the concern, (3) the specific assumption
at risk, and (4) suggested remediation steps. For example, weak instrument
warnings recommend weak-instrument-robust inference methods (Anderson-Rubin,
LIML) and reduced-form reporting.

Fundamental untestable assumptions---SUTVA, unconfoundedness, no
manipulation---are documented but cannot generate data-driven warnings.
The system recommends sensitivity analyses (e.g., \pkg{sensemakr}, E-values)
for selection-on-observables designs.

Future versions could extend this framework with LLM-generated warnings that
flag context-specific concerns (e.g., detecting specification search patterns)
and heuristic checks for common errors (e.g., sample size warnings for
subgroup analyses). These extensions require additional research and testing
to ensure warnings are accurate and actionable rather than overwhelming users
with false positives.

\subsection{LLM as method selector}

Large language models are the interpretive layer between user intent
and statistical execution. The LLM parses natural language requests, selects
appropriate methods from the toolkit, generates structured tool calls, and
interprets results. This builds on tool-augmented LLMs
\citep{Yao+etal:2023, Schick+etal:2023}.

Unlike code-generation approaches, our system does not ask the LLM to write
statistical code. Studies show LLMs generate incorrect code frequently
\citep{Nejjar+etal:2024}. Instead, we provide pre-implemented, verified methods
as callable tools. The LLM's constrained role (method selection and parameter
specification) enables smaller models to perform adequately (Section~\ref{sec:local}).

\begin{table}[htbp]
\centering
\caption{Comparison with existing LLM-based data analysis systems}
\label{tab:system-comparison}
\small
\begin{tabular}{lcccc}
\toprule
Feature & \pkg{prompt2analytics} & Econometrics AI & Data Interpreter & LAMBDA \\
\midrule
Approach & Tool selection & Enhanced code gen & Code gen + exec & Agent orchestration \\
Computation & Pre-validated Rust & LLM-generated & LLM-generated & LLM-generated \\
Local execution & Yes (Ollama) & No & No & No \\
Correctness & Validated vs R & Code review & None & None \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:system-comparison} contrasts \pkg{prompt2analytics} with related
systems \citep{Chen+Shen+Ma:2025, Hong+etal:2024, Sun+etal:2025:LAMBDA}---all of
which generate and execute code, whereas our system routes requests to
pre-validated implementations.

\subsection{Why Rust}

\proglang{Rust} compiles to native code with \proglang{C}-level performance
while guaranteeing memory safety at compile time, eliminating buffer overflows,
use-after-free, and data races that can cause incorrect results in numerical
software. The ecosystem provides \pkg{polars} for data manipulation,
\pkg{ndarray} for arrays, and \pkg{faer} for linear algebra. A single
codebase compiles to standalone binaries, WebAssembly for browsers, and
shared libraries for language integration.

\subsection{MCP as the bridge}

The Model Context Protocol (MCP) provides the communication layer between
LLMs and the analytics engine \citep{MCP:2024}.%
\footnote{MCP focuses on model-to-tool integration. Complementary protocols
address agent-to-agent communication: Google's Agent2Agent (A2A) protocol
\citep{A2A:2025} enables discovery and task delegation between AI agents,
while IBM's Agent Communication Protocol (ACP) \citep{ACP:2025} provides
lightweight HTTP-native messaging. A2A and ACP merged under Linux Foundation
governance in 2025. For single-agent tool invocation, the pattern used in
\pkg{prompt2analytics}, MCP remains the dominant standard, with adoption
by OpenAI, Microsoft, and Google alongside Anthropic.}
MCP defines a standardized way for language models to discover available
tools, invoke them with structured parameters, and receive structured results.

Each statistical method is exposed as an MCP tool with a defined schema
specifying required and optional parameters. For example, the OLS tool
accepts a dataset identifier, dependent variable, independent variables,
and options for robust standard errors. The LLM generates tool calls
conforming to this schema, and the MCP server routes them to the
appropriate \proglang{Rust} implementation.

This architecture decouples the language understanding layer from the
computation layer. The LLM can be swapped (Claude, GPT, Ollama) without
changing the analytics engine. Conversely, the statistical implementations
can be updated or extended without modifying the LLM integration.

\subsection{Data locality and privacy}

Bulk data processing occurs locally.
When a dataset is loaded, only metadata is transmitted to the LLM:
column names, data types, row counts, and null percentages. The LLM
uses this schema information to formulate appropriate tool calls.
Statistical computations (matrix operations, parameter estimation,
inference) execute entirely on the local machine, with only aggregated
results (coefficients, standard errors, test statistics) returned
through the LLM.

\subsubsection{Tools that transmit data}

Users should understand which operations may transmit data beyond
metadata and aggregated results:

\begin{itemize}
  \item \emph{Data preview} (\code{head}, \code{tail}, \code{sample}):
    Returns actual row values for inspection. Requested sample size
    determines transmission volume.
  \item \emph{Data profiling}: May include frequent values from categorical
    columns, min/max values for numeric columns, and example values for
    pattern detection.
  \item \emph{Outlier detection}: Reports specific observations identified
    as outliers, including their variable values.
  \item \emph{Error messages}: Failed operations may return the problematic
    values in error context (e.g., ``column `salary' contains non-numeric
    value `N/A' at row 47'').
\end{itemize}

Statistical estimation tools---regression, panel methods, hypothesis
tests---transmit only summary statistics (coefficients, standard errors,
test statistics, $p$-values), not individual observations. Residual
diagnostics report aggregate properties (normality tests, autocorrelation)
rather than observation-level residuals unless specifically requested.

\subsubsection{Information leakage through aggregates}

Even aggregated statistics can leak sensitive information in certain
contexts. Coefficient estimates in small-sample regressions may be
influenced by individual observations. Group means with small group
sizes (particularly singletons) directly reveal individual values.
Summary statistics for highly skewed distributions may identify
outlying individuals.

Users working with sensitive data should consider whether aggregated
output could enable re-identification or inference about protected
individuals. Local LLM deployment via Ollama eliminates these concerns
entirely by keeping all data and results on the local machine.

\subsubsection{Security considerations}

MCP implementations face security challenges common to tool-augmented
LLM systems. \emph{Prompt injection} attacks embed malicious instructions
in data that the LLM then processes, potentially causing unintended
tool invocations. \emph{Tool poisoning} attacks modify tool descriptions
to influence LLM behavior \citep{OWASP:MCP:2025}.

\pkg{prompt2analytics} mitigates these risks through several design
choices:
\begin{itemize}
  \item Tools have fixed, auditable descriptions embedded in the server
    binary---external modification is not possible without rebuilding.
  \item The server executes only pre-defined statistical methods.
    No arbitrary code execution capability exists; the LLM cannot
    instruct the server to run shell commands or access files outside
    loaded datasets.
  \item Tool calls are logged and can be inspected before execution
    (in interactive mode) or reviewed in audit logs.
  \item Dataset access requires explicit loading; the server does not
    have ambient filesystem access.
\end{itemize}

Nevertheless, users deploying MCP servers should follow security best
practices: avoid loading untrusted data files that could contain
injection payloads, review tool calls when processing sensitive data,
and prefer local LLM deployment for high-security environments.

The same interface and analytics engine function identically with
local or cloud-hosted models, allowing users to select the deployment
appropriate to their security requirements.

\subsection{Implemented methods}

The analytics engine implements over 200 econometric and statistical methods
organized into the following categories:

\begin{itemize}
  \item \emph{Regression}: OLS with HC0--HC3 robust standard errors, clustered
    standard errors, HAC (Newey-West), bootstrap inference, GLS, nonlinear
    least squares, LOESS, smoothing splines, stepwise selection, and
    quantile regression \citep{Koenker+Bassett:1978}.
  \item \emph{Panel data}: Fixed effects, random effects, Hausman test,
    high-dimensional fixed effects (HDFE), panel GLS, Arellano-Bond GMM
    for dynamic panels \citep{Arellano+Bond:1991}, and panel unit root tests
    (LLC, IPS, Fisher).
  \item \emph{Instrumental variables}: 2SLS with first-stage diagnostics,
    Sargan overidentification test, and general GMM \citep{Hansen:1982}.
  \item \emph{Causal inference---design-based}: Canonical difference-in-differences,
    staggered DiD via Callaway-Sant'Anna \citep{Callaway+SantAnna:2021},
    Goodman-Bacon decomposition \citep{Goodman-Bacon:2021}, extended TWFE
    \citep{Wooldridge:2021}, sharp and fuzzy regression discontinuity with
    CCT bandwidth selection \citep{Calonico+etal:2014}, multi-cutoff RD,
    synthetic control \citep{Abadie+etal:2010}, generalized synthetic control
    \citep{Xu:2017}, and synthetic control with prediction intervals \citep{Cattaneo+etal:2021}.
  \item \emph{Causal inference---treatment effects}: Inverse probability weighting,
    doubly robust/AIPW estimation, propensity score matching (nearest neighbor,
    CEM, full matching, subclassification) following the \pkg{MatchIt} interface
    \citep{Ho+etal:2011}, covariate balancing propensity scores \citep{Imai+Ratkovic:2014},
    entropy balancing, GBM-based weighting, TMLE \citep{vanderLaan+Rose:2011},
    collaborative TMLE, longitudinal TMLE, and double/debiased machine learning
    \citep{Chernozhukov+etal:2018}.
  \item \emph{Discrete choice}: Binary logit and probit, multinomial logit,
    conditional/McFadden logit, mixed logit with random parameters, ordered
    logit/probit, negative binomial, zero-inflated Poisson/negative binomial,
    and hurdle models.
  \item \emph{Time series}: ARIMA, VAR, VARMA, VECM, impulse response functions,
    Granger causality, seasonal decomposition (STL, MSTL), Holt-Winters,
    structural time series, Kalman filtering, GARCH, and changepoint detection.
  \item \emph{Spatial econometrics}: Spatial autoregressive (SAR), spatial error
    (SEM), and combined (SAC) models \citep{Anselin:1988}; spatial probit;
    spatial panel models; and spatial GMM with heteroskedasticity-robust
    inference.
  \item \emph{Survival analysis}: Kaplan-Meier estimation, log-rank tests,
    Cox proportional hazards, accelerated failure time models, and competing
    risks analysis.
  \item \emph{Mediation and sensitivity}: Causal mediation analysis, natural
    effect models, sensitivity analysis for unmeasured confounding
    \citep{Cinelli+Hazlett:2020}, E-values, and Balke-Pearl bounds.
  \item \emph{Statistical tests}: Over 50 hypothesis tests including $t$-tests,
    ANOVA, MANOVA, chi-squared, Fisher exact, Wilcoxon, Kruskal-Wallis,
    Friedman, normality tests (Shapiro-Wilk, Kolmogorov-Smirnov), and
    post-hoc procedures (Tukey HSD).
\end{itemize}

Full documentation of all methods, including mathematical specifications and
implementation details, is available in the package documentation and
Appendix~\ref{app:implementation}.

\subsection{Implementation caveats}

While method coverage is broad, several limitations remain relative to
state-of-the-art practice:
\begin{itemize}
  \item 2SLS reports first-stage $F$-statistics but lacks Stock-Yogo critical
    values for formal weak instrument inference.
  \item ARIMA supports non-seasonal and partially seasonal models; full
    seasonal ARIMA (SARIMA) with arbitrary seasonal periods is not yet
    implemented.
  \item Bayesian methods and Heckman selection models are not implemented;
    users requiring these should use \proglang{R} or \proglang{Stata}.
\end{itemize}

Appendix~\ref{app:implementation} provides detailed documentation of
method-specific limitations and recommended alternatives for edge cases.

\subsection{Best practices for effective use}

Realizing the benefits of chat-first analytics requires understanding when
the system excels and when traditional software remains preferable. This
section provides practical guidance based on our extended evaluation
(Section~\ref{sec:local}) and user testing.

\subsubsection{Pre-analysis checklist}

Before running an analysis through \pkg{prompt2analytics}, users should verify:
\begin{enumerate}
  \item \emph{Method availability}: Confirm the required method is implemented.
    Table~\ref{tab:feature-comparison} in Section~\ref{sec:comparison} provides
    feature coverage; Appendix~\ref{app:implementation} gives comprehensive
    method documentation.
  \item \emph{Data requirements}: Verify the dataset meets method assumptions.
    Panel methods require entity and time identifiers; IV requires instruments;
    DiD requires treatment and time indicators.
  \item \emph{Variable naming}: Use clear, unambiguous column names. The LLM
    extracts variable names from natural language; names like \code{x1} or
    \code{var\_1} are more error-prone than \code{wage} or \code{education}.
  \item \emph{Specification clarity}: Plan to specify methods explicitly when
    precision matters. ``Estimate a within-transformation fixed effects model
    with clustered standard errors at the firm level'' yields more reliable
    results than ``run a regression with firm effects.''
\end{enumerate}

\subsubsection{During analysis}

Multi-turn accuracy drops to 50--60\% (Section~\ref{sec:local}), so users should:
\begin{itemize}
  \item \emph{Verify tool selection}: Check which method was invoked before
    interpreting results. The system reports tool calls explicitly.
  \item \emph{Inspect extracted parameters}: Confirm that variable names,
    options (robust SEs, clustering), and specifications match intent.
  \item \emph{Expect corrections}: Budget for 2--3 clarification turns per
    analytical workflow. Rephrase ambiguous requests with explicit method names.
  \item \emph{Cross-check novel results}: For high-stakes analyses, verify
    key findings against \proglang{R} or \proglang{Stata} implementations.
\end{itemize}

\subsubsection{When to use traditional software}

\pkg{prompt2analytics} complements rather than replaces established tools.
Users should prefer \proglang{R}, \proglang{Stata}, or \proglang{Python} when:
\begin{itemize}
  \item \emph{Methods are unavailable}: Bayesian inference, Heckman selection,
    SARIMA with arbitrary seasonal periods, and Conley spatial SEs require
    specialized packages.
  \item \emph{Maximum performance is critical}: \pkg{fixest} outperforms
    \pkg{prompt2analytics} for high-dimensional fixed effects at scale.
    Simulation studies with thousands of replications may benefit from
    direct code.
  \item \emph{Reproducibility requires code}: Published research should
    include reproducible scripts. While \pkg{prompt2analytics} logs tool
    calls, traditional scripts provide clearer audit trails.
  \item \emph{Complex data pipelines}: Multi-step transformations with
    intermediate validations are more reliably implemented in scripted
    workflows.
  \item \emph{Visualization customization}: While \pkg{prompt2analytics}
    generates standard plots, publication-quality figures often require
    fine-grained control available in \pkg{ggplot2} or \pkg{matplotlib}.
\end{itemize}

\subsubsection{Recommended workflows}

The system is most effective for:
\begin{itemize}
  \item \emph{Exploratory analysis}: Quickly testing specifications before
    committing to a formal analysis pipeline.
  \item \emph{Teaching and learning}: Students can explore econometric methods
    conversationally while seeing which tools and parameters are invoked.
  \item \emph{Rapid prototyping}: Researchers can iterate on model specifications
    in natural language before writing production code.
  \item \emph{Verification}: Cross-checking results from other software by
    running the same analysis through an independent implementation.
  \item \emph{Accessibility}: Users with limited programming experience can
    access validated statistical methods without learning specialized syntax.
\end{itemize}

