\documentclass[11pt,a4paper]{article}

%% -- LaTeX packages -----------------------------------------------------------

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

%% Page layout
\usepackage[margin=1in]{geometry}

%% Math and symbols
\usepackage{amsmath,amssymb}
\usepackage{bm}

%% Tables and figures
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{graphicx}

%% Bibliography
\usepackage{natbib}
\bibliographystyle{plainnat}

%% Hyperlinks (load last)
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

%% -- Custom commands (compatibility with JSS) ---------------------------------

%% Package/software names
\newcommand{\pkg}[1]{\textbf{#1}}
\newcommand{\proglang}[1]{\textsf{#1}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}

%% Checkmark for feature comparison tables
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

%% Code/output environments (JSS-compatible)
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{Code}{Verbatim}{fontsize=\small,frame=single}
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontsize=\small,formatcom=\color{blue}}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{fontsize=\small}
\usepackage{xcolor}

%% Auto-generated output inclusion (styled like Soutput)
\newcommand{\CliOutput}[1]{\VerbatimInput[fontsize=\small]{#1}}


%% -- Title and authors --------------------------------------------------------

\title{Chat-First Data Analytics}

\author{
  First Author\thanks{Department of Economics, University of Somewhere.
  Email: \email{author@example.org}}
}

\date{\today}


\begin{document}

\maketitle

%% -- Abstract -----------------------------------------------------------------

\begin{abstract}
This paper introduces \emph{chat-first data analytics}: an approach where
natural language conversation is the primary interface to statistical
methods, while pre-implemented algorithms perform the actual computation.
Unlike approaches where large language models generate statistical code,
which raises concerns about correctness, data privacy, and reproducibility,
chat-first analytics separates interpretation (understanding user intent) from
computation (executing validated methods). The language model selects tools;
verified implementations produce results. We present \pkg{prompt2analytics},
a software implementation with an econometrics library in pure \proglang{Rust},
exposed through the Model Context Protocol (MCP) for integration with language
models. The library covers regression analysis with robust and clustered
standard errors, panel data models, instrumental variables,
difference-in-differences, discrete choice, and time series methods.
All implementations are validated against established \proglang{R} packages.
Local deployment via Ollama enables complete data isolation for sensitive
data analysis.
\end{abstract}

\medskip
\noindent\textbf{Keywords:} chat-first analytics, natural language interface, econometrics,
\proglang{Rust}, MCP, local LLM, data privacy

\bigskip

%% Include shared content
\input{paper}

%% =============================================================================
%% BIBLIOGRAPHY
%% =============================================================================

\bibliography{references}

\end{document}
