\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

%% additional packages
\usepackage{booktabs}
\usepackage{threeparttable}
%% Ensure table notes use footnotesize and no paragraph indent
\renewcommand{\TPTnoteSettings}{\footnotesize\setlength{\parindent}{0pt}}
\usepackage{multirow}
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{graphicx}

%% custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% Checkmark for feature comparison tables
\usepackage{pifont}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

%% Auto-generated output inclusion (styled like Soutput)
\newcommand{\CliOutput}[1]{\VerbatimInput[fontsize=\small]{#1}}

%% -- Article metainformation (author, title, ...) -----------------------------

\author{Ulrich Matter\\Bern University of Applied Sciences\\University of St.~Gallen}
\Plainauthor{Ulrich Matter}

%% Title
\title{Chat-First Data Analytics}
\Plaintitle{prompt2analytics: A Rust Toolkit for Econometrics and Data Analytics}
\Shorttitle{\pkg{prompt2analytics}: Econometrics in \proglang{Rust}}

%% Abstract
\Abstract{
  Using natural language to drive data analysis is an old idea, but large
  language models (LLMs) have made it suddenly ubiquitous. A widely observed
  and rather concerning workflow has users upload a dataset to ChatGPT
  and ask for an analysis; the model generates and executes code on a
  remote virtual machine, producing results that are difficult to
  verify, difficult to reproduce, and incompatible with data-privacy
  requirements. This paper describes an alternative approach called
  \emph{chat-first data analytics}, built on three principles: separate
  interpretation from computation, pre-validate methods rather than
  generate code, and transmit only metadata---not data. The language model
  interprets intent and selects tools; pre-implemented, validated
  algorithms produce results.
  \pkg{prompt2analytics} implements this approach as an econometrics and
  data analytics library in pure \proglang{Rust}, exposed through the
  Model Context Protocol~(MCP) for integration with language models. The
  library covers regression, panel data, instrumental variables,
  difference-in-differences, discrete choice, time series, and over 200
  additional methods---all validated against established \proglang{R}
  packages. Benchmarks show a median speedup of 4$\times$ over
  corresponding \proglang{R} implementations at $n = 100{,}000$.
  Evaluation shows that tool selection achieves
  90--100\% accuracy on curated prompts. Local deployment via Ollama enables
  complete data isolation on consumer hardware.
}

%% Keywords (at least one required)
\Keywords{chat-first analytics, natural language interface, econometrics,
  \proglang{Rust}, MCP, local LLM, data privacy}
\Plainkeywords{chat-first analytics, natural language interface, econometrics,
  Rust, Model Context Protocol, local LLM, data privacy}

\Address{
  Ulrich Matter\\
  Bern University of Applied Sciences, Bern, Switzerland\\
  University of St.~Gallen, St.~Gallen, Switzerland\\
  E-mail: \email{ulrich.matter@unisg.ch}\\
  URL: \url{https://umatter.github.io/}
}

\begin{document}

%% Include shared content
\input{paper}

%% =============================================================================
%% BIBLIOGRAPHY
%% =============================================================================

\bibliography{references}

\end{document}
